---
title: "Project Predictive Analytics"
author: "746 - Subhodeep Bhattacharjee"
date: "2026-05-02"
output: html_document
---

## Predicting **v_decile_score** using Criminal Offender Screening Data (Florida, 2013–2014)
#### This project focuses on building predictive models for the response variable: v_decile_score using demographic, criminal history, and custody-related predictors from a criminal offender screening dataset.

*Data Information* - The dataset is available as "crimOffenders_df" under the package "crimedatasets" package in R. \ 

**Variable Descrption** \
A data frame with 5,855 observations and 16 variables: \

age - Age of the offender (numeric) \

juv_fel_count - Number of juvenile felonies committed (numeric) \

decile_score - COMPAS score decile (numeric) \

juv_misd_count - Number of juvenile misdemeanors committed (numeric) \

juv_other_count - Number of other juvenile convictions (numeric) \

v_decile_score - Predicted decile score of the offender (numeric) \

priors_count - Number of prior crimes committed (numeric) \

sex - Gender of the offender (factor with levels 'Female' and 'Male') \

two_year_recid - Recidivism within two years (factor with levels 'Yes' and 'No') \

race - Race of the offender (factor with levels 'White', 'Black', 'Hispanic', 'Asian', 'Other', 'Native') \

c_jail_in - Date of entry into jail (normalized between 0 and 1, numeric) \ 

c_jail_out - Date of release from jail (normalized between 0 and 1, numeric) \

c_offense_date - Date the offense was committed (numeric) \

screening_date - Date the offender was screened (numeric) \

in_custody - Date the offender was placed in custody (numeric, normalized between 0 and 1) \

out_custody - Date the offender was released from custody (numeric, normalized between 0 and 1) \


/

/


##### Loading required libraries

```{r}
library(ISLR)
library(MASS)
library(AER)
library(crimedatasets)
library(carData)
library(car)
library(scatterplot3d)
library(Matrix)
library(leaps)
library(base)
library(glmnet)
library(dplyr)
library(tidyselect)
library(tidyr)
library(corrplot)
```


**Structure of the Dataset**
```{r}
attach(crimOffenders_df)
dim(crimOffenders_df)
head(crimOffenders_df)
summary(crimOffenders_df)
```
\

\

**Convert categorical variables into factors** 
```{r}
#renaming dataset
crime = crimOffenders_df

# converting the categorical variables to factor which are easy to model and interpret in regression
crime$sex = as.factor(crime$sex)
crime$race = as.factor(crime$race)
crime$two_year_recid = as.factor(crime$two_year_recid)
```

\

\

**Missing Value Detection and Handling the Missing values for the Dataset** \
```{r}
# Counting the number of missing values
colSums(is.na(crime))
```
Here we see that the dataset has no missing values. \


If there was any missing value in the dataset then we would have replaced it with median values of the respective column(variable) rather than omitting because in this case we do not miss the other important observation taken for different variables which can be of our use to perform analysis.

```{r}
# imputing with mean
numeric_cols = sapply(crime, is.numeric)

for(i in names(crime)[numeric_cols]){
  crime[[i]][is.na(crime[[i]])] = median(crime[[i]], na.rm = TRUE)
}
```


#### Exploratory Data Analysis (EDA)

```{r}
summary(crime)
```
**Interpretation** \
Mean shows average behavior \
Median helps identify skewness \
Large differences between mean and median indicate outliers \


**BoxPlot for Outlier detection**
```{r}
#taking numeric columns
numeric_data=select_if(crime,is.numeric)

par(mfrow=c(3,3))

for(i in names(numeric_data)){
  boxplot(numeric_data[[i]],
          main=i,
          col="lightblue")
}
```
Points outside whiskers indicate potential outliers and here we see that variables like priors_count may contain extreme outliers amongst others. \


**Correlation Matrix**
```{r}
cor_matrix = cor(numeric_data)

round(cor_matrix,2)

#plot showing the correlation among variables
corrplot(cor_matrix,
         method = "color",
         type = "upper",
         tl.cex = 0.8,
         number.cex = 0.7)
```
Correlation close to +1 → strong positive relationship \
Correlation close to -1 → strong negative relationship \
High correlation among predictors may create multicollinearity \

hence we pay special attention to: \

decile_score\
priors_count\
juvenile counts\
custody variables\

These are likely important predictors.\


### Now our main task is to optimized a model for predicting the decile score of the offender (v_decile_score) \

**We split the dataset into a Train and Test Data** \
Train Data or Training Sample is used build a model while Test Data/Sample is used to check the accuracy of the model built. \

```{r}
set.seed(123)
# dividing the data as 70% train data and rest 30% as test data

train_index = sample(0.7*nrow(crime),)

train = crime[train_index, ]
test = crime[-train_index, ]

dim(train)
dim(test)
```
##### Fitting a FULL Multiple Linear Regression Model
```{r}
full_model=lm(v_decile_score ~ . ,data=train)
summary(full_model)
```
**Interpretation** \
Here we see that in full model the variables age, juv_fel_count,jev_other_count, priors_count, sex, caucasian race and c_jail_out are important predictors amongst all. 


**Prediction of MSE on Train and Test Data for Full Model** \
```{r}
train_pred = predict(full_model, newdata = train)
test_pred = predict(full_model, newdata = test)

#MSE
train_mse = mean((train$v_decile_score - train_pred)^2)
train_mse
test_mse = mean((test$v_decile_score - test_pred)^2)
test_mse
```
The Test MSE is lesser than Train MSE so model is not a overfit. \


**Residual Plot** \
```{r}
plot(full_model,which=1)
```
Visible Outliers Detected. \

```{r}
# outliers
std_res = rstandard(full_model)
outliers = which(abs(std_res) > 3)  #I have considered the value of residual to be 3 rather than 2.5
outliers
```

**Levarage Point Plot** \
```{r}
plot(full_model, which = 5)
```

Leverage Points and Influential Point detected in graph \ 

```{r}
# Levarage Point
lev = hatvalues(full_model)
p = length(full_model$coefficients)
n = nrow(train)

cutoff = 3*(p+1)/n   #formula
cutoff
leverage_points = which(lev > cutoff)
# High leverage points may influence regression outcome
leverage_points
```



```{r}
# Influential Points
cook = cooks.distance(full_model)
plot(cook,
     type="h",
     main="Cook's Distance")

abline(h = 1,
       col="red")

#Formula
influential = which(cook > 1)
influential
```
No Influential points detected. \ 

**Removing these influencing observation from the sample** \
```{r}
remove_points = unique(c(outliers,
                          leverage_points,
                          influential))
train_clean = train[-remove_points, ]
```


**Refitting the full model after a clean train data** \
```{r}
full_model_clean = lm(v_decile_score ~ ., data = train_clean)

summary(full_model_clean)
```

```{r}
#comparison of both models
summary_full=summary(full_model)
summary_full$adj.r.squared

summary_full_clean=summary(full_model_clean)
summary_full_clean$adj.r.squared
```
Here we see that after removing the influential observations from the train data our model with new dataset performs better than the full model as the Adjusted R-Squared Value increases. Hence we will consider the cleaned full model data for further comparison and regressions. 


**Here in this Full Model all variables are used to predict the response, in further step of refinement we check for a subset where the same response can be predicted using less numbers of predictors.** \



#### Subset Selection for Model

**Fitting a subset using the forward stepwise technique**

```{r}
forward_model = regsubsets(
  v_decile_score ~ .,
  data = train,
  nvmax = 15,
  method = "forward"
)
forward_summary = summary(forward_model)
```

**Plot for Adjusted-R sqaure, Mallows Cp and BIC**  \
```{r}
#adjusted R squared
plot(forward_summary$adjr2,
     type = "b",
     xlab = "Number of Variables",
     ylab = "Adjusted R²")
```
\
Model with Higher Adjusted R$^2$ is preferred but we can see here that is uses all 15 variables in that case. \

```{r}
#Mallow's Cp Measure
plot(forward_summary$cp,
     type = "b",
     xlab = "Number of Variables",
     ylab = "Cp")
```
in Cp measure the lower value indicates better model also where the value is near the max number of predictors. \

```{r}
# BIC
plot(forward_summary$bic,
     type = "b",
     xlab = "Number of Variables",
     ylab = "BIC")
```
Since BIC penalizes complex model heavily so we preffer lower BIC value for model selection. \


**Finding the exact values of measures and identifying model** \
```{r}
which.max(forward_summary$adjr2)
which.min(forward_summary$cp)
which.min(forward_summary$bic)
```
building subsets model \

```{r}
coef(forward_model,which.max(forward_summary$adjr2))
coef(forward_model,which.min(forward_summary$cp))
coef(forward_model,which.min(forward_summary$bic))
```
```{r}
# best subset with variables out of 15
best_model_adjr2 = lm(v_decile_score ~ age + juv_fel_count + decile_score + juv_misd_count + juv_other_count + priors_count + sex + two_year_recid + race + c_jail_in + c_jail_out + c_offense_date + screening_date , data= train)
summary(best_model_adjr2)

best_model_cp = lm(v_decile_score ~ age + juv_fel_count + decile_score + juv_misd_count + juv_other_count + priors_count + sex + two_year_recid + race + c_jail_in + c_jail_out + screening_date , data = train )
summary(best_model_cp)

best_model_bic = lm(v_decile_score ~ age + decile_score + juv_misd_count + juv_other_count + priors_count + sex + race , data= train)
summary(best_model_bic)

```
Here model of adjr2 and cp has approximately equal R$^2$ and AdjR$^2$ value so we choose the best subset model with less number of predictors that is Cp which has 12 predictors. \


```{r}
best_model= lm(v_decile_score ~ age + juv_fel_count + decile_score + juv_misd_count + juv_other_count + priors_count + sex + two_year_recid + race + c_jail_in + c_jail_out + screening_date , data = train )
```


**Prediction of MSE for best subset model** \
```{r}
best_train_pred = predict(best_model,newdata = train)
best_test_pred = predict(best_model,newdata = test)

best_train_mse = mean((train_clean$v_decile_score - best_train_pred)^2)
best_train_mse

best_test_mse = mean((test$v_decile_score - best_test_pred)^2)
best_test_mse
```

**Comparing Full Model vs Best Subset Model** \
```{r}
comparison = data.frame(
  Model = c("Full Model", "Best Subset Model"),
  Train_MSE = c(train_mse, best_train_mse),
  Test_MSE = c(test_mse, best_test_mse)
)

comparison
```

**Here we prefer the model with lower test mse and lesser predictors so we see that Best Subset Model has lowest TEST MSE and also models the response with less number of predictors and higher Adjusted R$^2$ value.** \


\



**Since we have categorical variables in this dataset and multicollinearity then we perform the regression using Ridge and Lasso taking the values of Tuning Parameter through cross validation technique** \



#### Model through Ridge Regression 

Ridge helps reducing overfitting of model and multicollinearity among the variables. \


**Creating Model Matrix**

```{r}
x_train = model.matrix(v_decile_score ~ ., train)[,-1]
y_train = train$v_decile_score

x_test = model.matrix(v_decile_score ~ ., test)[,-1]
y_test = test$v_decile_score
```

**Cross Validation for Ridge for identifying Tuning Parameter** \

```{r}
set.seed(123)

ridge_cv = cv.glmnet(
  x_train,
  y_train,
  alpha = 0
)

plot(ridge_cv)
```
```{r}
# identifying the best Lambda - Tuning Parameter
ridge_cv$lambda.min
```

**Ridge Regression Model using Tuning Parameter** \
```{r}
#Ridge Model
ridge_model = glmnet(
  x_train,
  y_train,
  alpha = 0,
  lambda = ridge_cv$lambda.min
)
```
**Predicting Test MSE for Ridge** \
```{r}
ridge_pred = predict(ridge_model, s = ridge_cv$lambda.min, newx = x_test)
ridge_mse = mean((y_test - ridge_pred)^2)
ridge_mse
```
#### Model through Lasso

LASSO helps for variable selection and shrinkage of coefficients making model performance better. \

**Cross Validation for chosing the best Tuning Parameter for Lasso** \
```{r}
set.seed(123)

lasso_cv = cv.glmnet(
  x_train,
  y_train,
  alpha = 1
)
plot(lasso_cv)
```
```{r}
#best lambda- tuning parameter
lasso_cv$lambda.min
```

**Lasso model using Tuning Parameter** \
```{r}
#Lasso Model
lasso_model = glmnet(
  x_train,
  y_train,
  alpha = 1,
  lambda = lasso_cv$lambda.min
)
```

```{r}
#coefficients of Lasso Model
coef(lasso_model)
```

Here we see that variables with 0 regression coefficients have been removed. \

**Predicting MSE for Lasso** \
```{r}
lasso_pred = predict(lasso_model, s = lasso_cv$lambda.min, newx = x_test)
lasso_mse = mean((y_test - lasso_pred)^2)
lasso_mse
```

### Model Comparison 
```{r}
final_results = data.frame(
  Model = c(
    "Full Regression",
    "Best Subset",
    "Ridge",
    "Lasso"
  ),
  Test_MSE = c(
    test_mse,
    best_test_mse,
    ridge_mse,
    lasso_mse
  )
)

final_results
```

**Conclusion** \
Since the TEST MSE of *Best subset* selected through forward stepwise selection is lower we choose this model over others, here the number of predictors  used is also less as compared to other model. \
Important predictors are age, COMPASS score decile, Number of other juvenile convictions, Race, Number of prior crimes committed and recidivism within two years as these directly reflect criminal history and risk. \


###### Findings
Criminal history variables strongly influence violent decile risk score. \
Prior offenses and recidivism indicators are major predictors. \
Ridge and Lasso improve prediction stability. \
Removing influential observations improves model reliability. \
Forward subset selection reduces unnecessary variables. \